---
title: "A Bioconductor-Native, Composable Heatmap"
author: "SummarizedHeatmap developers"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A Bioconductor-Native, Composable Heatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE,
    dev = "png",
    fig_crop = FALSE,
    fig.width = 7,
    fig.height = 5
)
options(bitmapType = "cairo")
library(SummarizedHeatmap)
```

# Motivation

Heatmaps are often built from a matrix and a collection of separate annotation
vectors. Keeping those pieces synchronized becomes difficult when data are
subsetted, reordered, or reused in a downstream analysis. `SummarizedHeatmap`
stores the assay, `rowData`, `colData`, display order, and any hierarchical
clusterings in one `SummarizedExperiment`-derived object. It separates that
data representation from rendering: each visual component is a regular
ggplot2 plot, and patchwork handles layout and composition.

This makes the package a natural fit for Bioconductor workflows, where assays
and their feature and sample metadata are already represented by
`SummarizedExperiment`. It also lets users continue to work with familiar
ggplot2 scales, themes, and geoms after making a heatmap component.

This approach complements other heatmap packages. [ComplexHeatmap](https://bioconductor.org/packages/ComplexHeatmap/)
provides a comprehensive system for arranging heatmaps and annotations. The
CRAN package [ggheatmapper](https://csgroen.github.io/ggheatmapper/) also uses
ggplot2 and patchwork for tile-based heatmaps. `SummarizedHeatmap` focuses on a
different boundary: persistent assay, metadata, order, and clustering state
inside a Bioconductor class, with individually accessible ggplot components
for users who want to compose their own layouts.

# Installation

For a released version, install with BiocManager:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("SummarizedHeatmap")
```

# Create a heatmap object

The small deterministic example below resembles a normalized gene-expression
assay. Feature and sample annotations are passed using the usual
`SummarizedExperiment` metadata classes.

```{r create}
set.seed(2026)
mat <- matrix(
    rnorm(48),
    nrow = 8,
    dimnames = list(paste0("gene", seq_len(8)), paste0("sample", seq_len(6)))
)
row_info <- S4Vectors::DataFrame(
    feature_class = rep(c("marker", "background"), each = 4),
    row.names = rownames(mat)
)
column_info <- S4Vectors::DataFrame(
    condition = rep(c("control", "treated"), each = 3),
    batch = rep(c("batch 1", "batch 2"), 3),
    row.names = colnames(mat)
)

hm <- SummarizedHeatmap(
    mat,
    rowData = row_info,
    colData = column_info
)
hm
```

Rows and columns are clustered independently using Euclidean distance and
average linkage by default. `rowOrder()` and `colOrder()` return integer
indices in display order; the corresponding dendrograms are available through
`rowDendro()` and `colDendro()`.

```{r state}
rowOrder(hm)
colOrder(hm)
class(rowDendro(hm))
class(colDendro(hm))
```

Annotations remain ordinary `rowData` and `colData` and can be read or
modified through the inherited SummarizedExperiment accessors.

# Plot components

`plotHeatmapMain()` returns a regular ggplot. Annotation functions create one
ggplot per variable; selecting multiple variables returns a patchwork of the
independent plots. Dendrogram functions also return ggplots.

```{r components}
main <- plotHeatmapMain(hm) +
    ggplot2::labs(fill = "Scaled expression") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
column_annotation <- plotColData(hm, vars = c("Condition" = "condition"))
row_annotation <- plotRowData(hm, vars = c("Feature class" = "feature_class"))
column_tree <- plotColDendro(hm)
row_tree <- plotRowDendro(hm)

class(main)
class(column_annotation)
```

`plotHeatmap()` provides a complete starting layout with annotations,
dendrograms, and a shared guide area. Column annotations are stacked vertically;
on the left of the heatmap, the row dendrogram is followed by row annotations,
so the annotations sit beside the heatmap body. The row dendrogram's leaves
face the heatmap. The component functions remain available when a custom layout
is preferred.

```{r complete}
complete <- plotHeatmap(hm)
complete
```

# Compose and customize with patchwork

The components can be arranged with patchwork operators, `wrap_plots()`, and
`plot_layout()`. `free()` releases a panel's alignment on selected sides, and
`guide_area()` reserves a location for guides when `guides = "collect"` is
used. In `plotHeatmap()`, the package constructs panels in a fixed grid
traversal order so collected guides are predictable: column dendrogram,
column annotations, row dendrogram, row annotations, then the main heatmap.

```{r custom-layout}
stacked <- patchwork::wrap_plots(
    list(column_tree, main),
    ncol = 1, heights = c(1, 5), guides = "collect"
)
stacked
```

```{r free-layout}
custom <- (column_annotation / main) +
    patchwork::plot_layout(heights = c(1, 5), guides = "collect")
custom
```

Arbitrary ggplots can be combined with heatmap components using the same
patchwork operations. Here a small sample-level summary plot sits beside the
heatmap.

```{r extra-plot}
display_samples <- colnames(mat)[colOrder(hm)]
qc <- ggplot2::ggplot(
    data.frame(
        sample = factor(display_samples, levels = display_samples),
        mean = colMeans(mat)[display_samples]
    ),
    ggplot2::aes(x = sample, y = mean)
) +
    ggplot2::geom_point() +
    ggplot2::labs(x = NULL, y = "Mean assay value") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

(qc / main) + patchwork::plot_layout(heights = c(1, 5))
```

# Subsetting and reclustering

Subsetting follows SummarizedExperiment semantics. If selected rows differ in
membership or order, the stored row dendrogram is cleared and the row order is
reset to the natural order of the subset. The column state is preserved when
columns are unchanged. Column subsetting applies the same rule independently.
Subsetting does not silently recluster; call a clustering helper explicitly
when a new dendrogram is wanted.

```{r subset}
hm_subset <- hm[1:5, c("sample1", "sample3", "sample5")]
validObject(hm_subset)
is.null(rowDendro(hm_subset))
is.null(colDendro(hm_subset))
rowOrder(hm_subset)
colOrder(hm_subset)
```

The retained columns still carry their previous clustering because their
membership and order did not change in this example. Recluster the subset
explicitly to compute clustering against the smaller assay:

```{r recluster}
hm_subset <- clusterHeatmap(hm_subset)
validObject(hm_subset)
```

# Session information

```{r session-info}
sessionInfo()
```
